Demo arXiv template

### 2023-01-30 paper/paper.qmd
first draft on HDI index

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, echo = FALSE)
library(tidyverse)
library(latex2exp)
library(patchwork)
```

@@

Weighting

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
options(width = 90)
library(tidyverse)
library(latex2exp)
library(patchwork)
library(indri)
library(GGally)
```

@@

<!-- Indexes are commonly used in social science is to rank entities (i.e. countries) on a topic through a range of indicators. This includes university ranking from QS, Times Higher Education; the Sustainable Development Goals (SDG) Index [@SDSN_Bertelsmann_2021] by the Sustainable Development Solutions Network (SDSN), the Human Development Index (HDI) [@UNDP_2021] by United Nations Development Programme (UNDP), etc. These indexes, or rankings, typically have a set of predefined dimensions, or pillars, and indicators are first reduced to these pillars and then reduced to an index.  -->

In the following example, we will first constructing the Human Development Index (HDI) using the pipeline syntax and then show how the index, and ranking, may changes once an alternative expression or parameter in the computation is changed in two scenarios.

Human Development Index (HDI) measures the development of countries from human living. Rather than focusing solely on the contribution of economic growth to the standard of living, HDI identifies another two important dimension in human development: a long living life and being knowledgeable. The three dimensions are measured using four indicators: life expectancy at birth (health), expected years of schooling (education), mean years of schooling (education), and Gross National Income (GNI) *per capita* (standard of living). The technical notes [@UNDP_2021_tech_notes] have documented the procedures to calculate HDI and they are summarised below:

  1. take log on GNI *per capita*,
  2. rescale the four indicators into [0, 1] using mini-max,
  3. aggregate the two education indicators using arithmetic mean
  4. aggregate the three dimensions into the index using geometric mean

The values used in mini-max rescaling is summarised in @tbl-rescale-params and the justification of these numbers can be found in the technical notes mentioned above.

```{r}
#| tbl-cap-location: bottom
#| tbl-cap: Maximum and minimum values used to rescale the four HDI indicators into [0, 1] range. The maximum of GNI per capita is taken as the common log of 75,000, approximatly 4.875.
#| label: tbl-rescale-params
scaling_params <- tibble::tribble(
   ~Dimension, ~Indicator, ~Variable,  ~Minimum, ~Maximum,
  "Health",              "Life expectancy at birth (years)",   "life_exp",    "20",          "85",
  "Education",           "Expected years of schooling (years)",  "exp_sch",    "0",          "18",
  "Education",           "Mean years of schooling (years)",   "avg_sch",      "0",          "15",
  "Standard of living",  "GNI per capita (2017 PPP$)",     "gni_pc",       "100",      "75000"
  ) %>%
  mutate(across(contains("mum"), as.numeric),
         across(contains("mum"), ~ifelse(Variable == "gni_pc", log10(.x), .x)))
knitr::kable(scaling_params, digits = 3)
```

Among the four steps listed in the calculation, the first two are variable transformation. The next two steps can be considered as dimension reduction given it intends to combine multivariate data into univariate. With the index pipeline proposed in the paper, HDI can be calculated as:

```{r}
dt <- readxl::read_xlsx(path = here::here("data/hdi.xlsx") , skip = 4)
raw <- dt %>%
  janitor::clean_names() %>%
  dplyr::select(-paste0("x", seq(4, 14, 2))) %>%
  dplyr::rename(id = x1, countries = x2) %>%
  dplyr::mutate(across(c(id, human_development_index_hdi:hdi_rank), as.numeric)) %>%
  dplyr::filter(!is.na(id)) %>%
  dplyr::mutate(across(4:7, ~round(.x, digits = 3))) %>% 
  dplyr::select(-c(human_development_index_hdi,
                   gni_per_capita_rank_minus_hdi_rank, 
                   hdi_rank)) 
colnames(raw) <- c("id", "country", "life_exp", "exp_sch", "avg_sch", "gni_pc")
```

```{r echo = TRUE, message=TRUE}
dt <- raw %>% init(id = country, indicators = life_exp:gni_pc)
(res <- dt %>%
  var_trans(gni_pc = log10(gni_pc)) %>% 
  var_trans(.method = rescale_minmax, .vars = life_exp:gni_pc,
            min = c(20, 0, 0, 2), max = c(85, 18, 15, log10(75000))) %>% 
  dim_red(sch = (exp_sch + avg_sch) / 2) %>%
  dim_red(index = (life_exp * sch * gni_pc)^(1/3)))
```

The output is an `indri` object with a summary of operations followed by the data output. The data output contains the ID, country name, rescaled variables (`life_exp`, `exp_sch`, `avg_sch`, and `gni_pc`), the aggregated education dimension (`sch`), and the final index calculated (`index`).

Index analysts can be interested in studying the property of the index and the countries through the "what-if" questions. In the context of HDI, this can be:

  - what if arithmetic mean, or different weights, is used to reduce the three dimensions into HDI? 
  - what if the maximum and minimum values used in the rescaling changes?

We will now demonstrate how these questions can be answered using the function `swap_exprs()` and `swap_values()`. Both function first use a set of arguments (`.var`, `.module`, `.step`, and `.res`) to locate the operation to be tested. Then, `.expr` and `.values` will accept a list of expression or a vector of values respectively for testing. Lastly, the initial `indri` (raw data) needs to be provided. The following code tests on five expressions (arithmetic mean, three weighting schemes each emphasizing on one dimension, and principal component analysis recommended weight):

```{r echo = TRUE, message=TRUE}
(res2 <- res %>%
  swap_exprs(
    .var = index,
    .expr = list(
      index1 = (life_exp + sch + gni_pc)/3,
      # PCA recommended weight
      index2 = 0.569 * life_exp + 0.576 * sch + 0.586 * gni_pc, 
      index3 = 0.8 * life_exp + 0.1 * sch + 0.1 * gni_pc,
      index4 = 0.1 * life_exp + 0.8 * sch + 0.1 * gni_pc,
      index5 = 0.1 * life_exp + 0.1 * sch + 0.8 * gni_pc),
    .raw_data = dt))
```

The resulting data includes five more columns corresponding to the five alternative index. The country ranking can be further computed on the data and plotted in a scatterplot matrix as in @fig-hdi-expr. The plot shows almost similar ranks suggested by the indexes calculated using geometric mean (`rank0`), the arithmetic mean (`rank1`), and the PCA recommended weights (`rank2`). The agreement of the three methods supports the use of geometric mean as the dimension reduction method in the original index. The scatterplot matrix also reveals the difference in ranking when the three dimensions are given different weights in `rank3`, `rank4`, and `rank5`. The panel `rank4` vs `rank5` shows a cluster of highlighted points with relatively high ranks according to `rank5` but low in `rank4`.  These countries are printed in @tbl-hdi-expr prints along with their three dimensions, calculated indexes and rankings. While having a high life expectancy and GNI *per capita*, these countries (Andorra, Qatar, San Marino, Kuwait, and Brunei Darussalam) score relatively low in education, which is weighted heavily in `index4` (`0.1 * life_exp + 0.1 * sch + 0.8 * gni_pc`).

```{r fig-hdi-expr}
#| fig-height: 8
#| fig-width: 12
#| fig-cap: "Scatterplot matrix of country rankings computed from different dimension reduction expressions in HDI: geometric mean (`rank0`), arithmetic mean (`rank1`), principal component analysis recommended weights (`rank2`), and heavier weight on life expectancy, education, and GNI per capita (`rank3` - `rank5`). Geometric mean, arithmetic mean, and PCA produce similar rankings where countries are aligned in a diagnoal line, while country rankings varies when compare among `rank3` to `rank5`. Panel `rank5` vs. `rank4` highlights a cluster of countries that rank high by `rank5` but low by `rank4`. The plot demonstrates the necessity to tests on various alternative expressions to study the robustness of the method and to capture distinctive characteristics of different countries. "
with_rank <- res2$data %>% 
  mutate(across(contains("index"), ~rank(-.x), .names = "rank.{.col}")) %>% 
  rename_with(~gsub(".index", "", .x), contains("rank.index"))

lower_cluster <- with_rank %>% 
  filter(rank5 < 30, rank4 > 50) %>% 
  select(country, life_exp, sch, gni_pc, index3, index4, rank3, rank4)

with_highlight <- with_rank %>% 
  mutate(highlight = ifelse(country %in% lower_cluster$country, TRUE, FALSE))

p1 <- ggpairs(
  with_highlight, columns = 14:19,
  lower = list(continuous = "points", combo = "facethist", 
               discrete = "facetbar", na = "na"),
  upper = NULL, diag = NULL, switch = "both") +
  theme(aspect.ratio = 1)

# https://stackoverflow.com/questions/65091593/how-to-mark-certain-point-in-ggpairs
p1$plots[[35]]$mapping <- 
    `class<-`(c(p1$plots[[35]]$mapping, aes(color = highlight)), "uneval")
p1 <- p1 + scale_color_manual(values = c("black", "red"))
p1
```

```{r echo = FALSE}
#| tbl-cap: A selected number of countries with low rankings when education is given a heavy weight (`index4`) while ranks high when a heavier weight is given on GNI per capital (`index5`).
#| label: tbl-hdi-expr
lower_cluster %>% 
  rename( `life expectancy` = life_exp, education = sch, `GNI per capita` = gni_pc) %>% 
  knitr::kable(digits = 3)
```

The maximum and minimum limit is carefully chosen with justification provided in the technical notes:

  > The low minimum value for gross national income (GNI) per capita, $100, is justified by the considerable amount of unmeasured subsistence and nonmarket production in economies close to the minimum, which is not captured in the official data

  What if this lower minimum is raised or decreased?

```{r echo = TRUE}
res3 <- res %>%
  swap_values(
    .module = var_trans, .step = rescale_minmax, .res = gni_pc,
    .var = min, .values = log10(700),
    .raw_data = dt)
```

### 2023-01-30 paper/paper.tex
first draft on HDI index

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, frame hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}, enhanced, interior hidden, boxrule=0pt]}{\end{tcolorbox}}\fi

@@

Weighting

@@

\leavevmode\vadjust pre{\hypertarget{ref-hdi}{}}%
United Nations Development Programme. 2022. {``Human Development Report
2021-22.''} New York. \url{http://report.hdr.undp.org}.

>>>>

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, frame hidden, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, enhanced, breakable, sharp corners]}{\end{tcolorbox}}\fi

@@

In the following example, we will first constructing the Human
Development Index (HDI) using the pipeline syntax and then show how the
index, and ranking, may changes once an alternative expression or
parameter in the computation is changed in two scenarios.

Human Development Index (HDI) measures the development of countries from
human living. Rather than focusing solely on the contribution of
economic growth to the standard of living, HDI identifies another two
important dimension in human development: a long living life and being
knowledgeable. The three dimensions are measured using four indicators:
life expectancy at birth (health), expected years of schooling
(education), mean years of schooling (education), and Gross National
Income (GNI) \emph{per capita} (standard of living). The technical notes
(United Nations Development Programme 2021) have documented the
procedures to calculate HDI and they are summarised below:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  take log on GNI \emph{per capita},
\item
  rescale the four indicators into {[}0, 1{]} using mini-max,
\item
  aggregate the two education indicators using arithmetic mean
\item
  aggregate the three dimensions into the index using geometric mean
\end{enumerate}

The values used in mini-max rescaling is summarised in
Table~\ref{tbl-rescale-params} and the justification of these numbers
can be found in the technical notes mentioned above.

\hypertarget{tbl-rescale-params}{}
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2375}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.4500}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1125}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1000}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1000}}@{}}
\caption{\label{tbl-rescale-params}Maximum and minimum values used to
rescale the four HDI indicators into {[}0, 1{]} range. The maximum of
GNI per capita is taken as the common log of 75,000, approximatly
4.875.}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Dimension
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Indicator
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Variable
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
Minimum
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
Maximum
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Dimension
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Indicator
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Variable
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
Minimum
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
Maximum
\end{minipage} \\
\midrule()
\endhead
Health & Life expectancy at birth (years) & life\_exp & 20 & 85.000 \\
Education & Expected years of schooling (years) & exp\_sch & 0 &
18.000 \\
Education & Mean years of schooling (years) & avg\_sch & 0 & 15.000 \\
Standard of living & GNI per capita (2017 PPP\$) & gni\_pc & 2 &
4.875 \\
\bottomrule()
\end{longtable}

Among the four steps listed in the calculation, the first two are
variable transformation. The next two steps can be considered as
dimension reduction given it intends to combine multivariate data into
univariate. With the index pipeline proposed in the paper, HDI can be
calculated as:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{dt }\OtherTok{\textless{}{-}}\NormalTok{ raw }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{init}\NormalTok{(}\AttributeTok{id =}\NormalTok{ country, }\AttributeTok{indicators =}\NormalTok{ life\_exp}\SpecialCharTok{:}\NormalTok{gni\_pc)}
\NormalTok{(res }\OtherTok{\textless{}{-}}\NormalTok{ dt }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{var\_trans}\NormalTok{(}\AttributeTok{gni\_pc =} \FunctionTok{log10}\NormalTok{(gni\_pc)) }\SpecialCharTok{\%\textgreater{}\%} 
  \FunctionTok{var\_trans}\NormalTok{(}\AttributeTok{.method =}\NormalTok{ rescale\_minmax, }\AttributeTok{.vars =}\NormalTok{ life\_exp}\SpecialCharTok{:}\NormalTok{gni\_pc,}
            \AttributeTok{min =} \FunctionTok{c}\NormalTok{(}\DecValTok{20}\NormalTok{, }\DecValTok{0}\NormalTok{, }\DecValTok{0}\NormalTok{, }\DecValTok{2}\NormalTok{), }\AttributeTok{max =} \FunctionTok{c}\NormalTok{(}\DecValTok{85}\NormalTok{, }\DecValTok{18}\NormalTok{, }\DecValTok{15}\NormalTok{, }\FunctionTok{log10}\NormalTok{(}\DecValTok{75000}\NormalTok{))) }\SpecialCharTok{\%\textgreater{}\%} 
  \FunctionTok{dim\_red}\NormalTok{(}\AttributeTok{sch =}\NormalTok{ (exp\_sch }\SpecialCharTok{+}\NormalTok{ avg\_sch) }\SpecialCharTok{/} \DecValTok{2}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{dim\_red}\NormalTok{(}\AttributeTok{index =}\NormalTok{ (life\_exp }\SpecialCharTok{*}\NormalTok{ sch }\SpecialCharTok{*}\NormalTok{ gni\_pc)}\SpecialCharTok{\^{}}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\DecValTok{3}\NormalTok{)))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
Index pipeline:

Summary: 
\end{verbatim}

\begin{verbatim}
var_trans: `log10(gni_pc)` -> gni_pc
\end{verbatim}

\begin{verbatim}
var_trans: `rescale_minmax(var = life_exp, min = 20.000000, max = 85.000000)` -> life_exp
\end{verbatim}

\begin{verbatim}
var_trans: `rescale_minmax(var = exp_sch, min = 0.000000, max = 18.000000)` -> exp_sch
\end{verbatim}

\begin{verbatim}
var_trans: `rescale_minmax(var = avg_sch, min = 0.000000, max = 15.000000)` -> avg_sch
\end{verbatim}

\begin{verbatim}
var_trans: `rescale_minmax(var = gni_pc, min = 2.000000, max = 4.875061)` -> gni_pc
\end{verbatim}

\begin{verbatim}
dim_red: `(exp_sch + avg_sch)/2` -> sch
\end{verbatim}

\begin{verbatim}
dim_red: `(life_exp * sch * gni_pc)^(1/3)` -> index
\end{verbatim}

\begin{verbatim}

Data: 
# A tibble: 191 x 8
      id country                life_exp exp_sch avg_sch gni_pc   sch index
   <dbl> <chr>                     <dbl>   <dbl>   <dbl>  <dbl> <dbl> <dbl>
 1     1 Switzerland               0.984   0.917   0.924  0.983 0.920 0.962
 2     2 Norway                    0.973   1.01    0.867  0.978 0.939 0.963
 3     3 Iceland                   0.964   1.06    0.918  0.955 0.991 0.970
 4     4 Hong Kong, China (SAR)    1.01    0.960   0.815  0.973 0.887 0.954
 5     5 Australia                 0.993   1.17    0.848  0.936 1.01  0.979
 6     6 Denmark                   0.944   1.04    0.864  0.967 0.952 0.954
 7     7 Sweden                    0.969   1.08    0.841  0.952 0.960 0.960
 8     8 Ireland                   0.954   1.05    0.772  1.00  0.912 0.955
 9     9 Germany                   0.933   0.945   0.939  0.952 0.942 0.942
10    10 Netherlands               0.949   1.04    0.839  0.956 0.939 0.948
# ... with 181 more rows
\end{verbatim}

The output is an \texttt{indri} object with a summary of operations
followed by the data output. The data output contains the ID, country
name, rescaled variables (\texttt{life\_exp}, \texttt{exp\_sch},
\texttt{avg\_sch}, and \texttt{gni\_pc}), the aggregated education
dimension (\texttt{sch}), and the final index calculated
(\texttt{index}).

Index analysts can be interested in studying the property of the index
and the countries through the ``what-if'' questions. In the context of
HDI, this can be:

\begin{itemize}
\tightlist
\item
  what if arithmetic mean, or different weights, is used to reduce the
  three dimensions into HDI?
\item
  what if the maximum and minimum values used in the rescaling changes?
\end{itemize}

We will now demonstrate how these questions can be answered using the
function \texttt{swap\_exprs()} and \texttt{swap\_values()}. Both
function first use a set of arguments (\texttt{.var}, \texttt{.module},
\texttt{.step}, and \texttt{.res}) to locate the operation to be tested.
Then, \texttt{.expr} and \texttt{.values} will accept a list of
expression or a vector of values respectively for testing. Lastly, the
initial \texttt{indri} (raw data) needs to be provided. The following
code tests on five expressions (arithmetic mean, three weighting schemes
each emphasizing on one dimension, and principal component analysis
recommended weight):

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{(res2 }\OtherTok{\textless{}{-}}\NormalTok{ res }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{swap\_exprs}\NormalTok{(}
    \AttributeTok{.var =}\NormalTok{ index,}
    \AttributeTok{.expr =} \FunctionTok{list}\NormalTok{(}
      \AttributeTok{index1 =}\NormalTok{ (life\_exp }\SpecialCharTok{+}\NormalTok{ sch }\SpecialCharTok{+}\NormalTok{ gni\_pc)}\SpecialCharTok{/}\DecValTok{3}\NormalTok{,}
      \CommentTok{\# PCA recommended weight}
      \AttributeTok{index2 =} \FloatTok{0.569} \SpecialCharTok{*}\NormalTok{ life\_exp }\SpecialCharTok{+} \FloatTok{0.576} \SpecialCharTok{*}\NormalTok{ sch }\SpecialCharTok{+} \FloatTok{0.586} \SpecialCharTok{*}\NormalTok{ gni\_pc, }
      \AttributeTok{index3 =} \FloatTok{0.8} \SpecialCharTok{*}\NormalTok{ life\_exp }\SpecialCharTok{+} \FloatTok{0.1} \SpecialCharTok{*}\NormalTok{ sch }\SpecialCharTok{+} \FloatTok{0.1} \SpecialCharTok{*}\NormalTok{ gni\_pc,}
      \AttributeTok{index4 =} \FloatTok{0.1} \SpecialCharTok{*}\NormalTok{ life\_exp }\SpecialCharTok{+} \FloatTok{0.8} \SpecialCharTok{*}\NormalTok{ sch }\SpecialCharTok{+} \FloatTok{0.1} \SpecialCharTok{*}\NormalTok{ gni\_pc,}
      \AttributeTok{index5 =} \FloatTok{0.1} \SpecialCharTok{*}\NormalTok{ life\_exp }\SpecialCharTok{+} \FloatTok{0.1} \SpecialCharTok{*}\NormalTok{ sch }\SpecialCharTok{+} \FloatTok{0.8} \SpecialCharTok{*}\NormalTok{ gni\_pc),}
    \AttributeTok{.raw\_data =}\NormalTok{ dt))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
Index pipeline:

Summary: 
\end{verbatim}

\begin{verbatim}
var_trans: `log10(gni_pc)` -> gni_pc
\end{verbatim}

\begin{verbatim}
var_trans: `rescale_minmax(var = life_exp, min = 20, max = 85)` -> life_exp
\end{verbatim}

\begin{verbatim}
var_trans: `rescale_minmax(var = exp_sch, min = 0, max = 18)` -> exp_sch
\end{verbatim}

\begin{verbatim}
var_trans: `rescale_minmax(var = avg_sch, min = 0, max = 15)` -> avg_sch
\end{verbatim}

\begin{verbatim}
var_trans: `rescale_minmax(var = gni_pc, min = 2, max = 4.875061)` -> gni_pc
\end{verbatim}

\begin{verbatim}
dim_red: `(exp_sch + avg_sch)/2` -> sch
\end{verbatim}

\begin{verbatim}
dim_red: `(life_exp * sch * gni_pc)^(1/3)` -> index0
\end{verbatim}

\begin{verbatim}
dim_red: `(life_exp + sch + gni_pc)/3` -> index1
\end{verbatim}

\begin{verbatim}
dim_red: `0.569 * life_exp + 0.576 * sch + 0.586 * gni_pc` -> index2
\end{verbatim}

\begin{verbatim}
dim_red: `0.8 * life_exp + 0.1 * sch + 0.1 * gni_pc` -> index3
\end{verbatim}

\begin{verbatim}
dim_red: `0.1 * life_exp + 0.8 * sch + 0.1 * gni_pc` -> index4
\end{verbatim}

\begin{verbatim}
dim_red: `0.1 * life_exp + 0.1 * sch + 0.8 * gni_pc` -> index5
\end{verbatim}

\begin{verbatim}

Data: 
# A tibble: 191 x 13
      id country   life_~1 exp_sch avg_sch gni_pc   sch index0 index1 index2 index3 index4
   <dbl> <chr>       <dbl>   <dbl>   <dbl>  <dbl> <dbl>  <dbl>  <dbl>  <dbl>  <dbl>  <dbl>
 1     1 Switzerl~   0.984   0.917   0.924  0.983 0.920  0.962  0.963   1.67  0.978  0.933
 2     2 Norway      0.973   1.01    0.867  0.978 0.939  0.963  0.963   1.67  0.970  0.946
 3     3 Iceland     0.964   1.06    0.918  0.955 0.991  0.970  0.970   1.68  0.966  0.985
 4     4 Hong Kon~   1.01    0.960   0.815  0.973 0.887  0.954  0.956   1.65  0.992  0.908
 5     5 Australia   0.993   1.17    0.848  0.936 1.01   0.979  0.979   1.69  0.989  1.00 
 6     6 Denmark     0.944   1.04    0.864  0.967 0.952  0.954  0.954   1.65  0.947  0.953
 7     7 Sweden      0.969   1.08    0.841  0.952 0.960  0.960  0.960   1.66  0.966  0.960
 8     8 Ireland     0.954   1.05    0.772  1.00  0.912  0.955  0.956   1.66  0.955  0.925
 9     9 Germany     0.933   0.945   0.939  0.952 0.942  0.942  0.942   1.63  0.936  0.942
10    10 Netherla~   0.949   1.04    0.839  0.956 0.939  0.948  0.948   1.64  0.949  0.941
# ... with 181 more rows, 1 more variable: index5 <dbl>, and abbreviated variable name
#   1: life_exp
\end{verbatim}

The resulting data includes five more columns corresponding to the five
alternative index. The country ranking can be further computed on the
data and plotted in a scatterplot matrix as in
Figure~\ref{fig-hdi-expr}. The plot shows almost similar ranks suggested
by the indexes calculated using geometric mean (\texttt{rank0}), the
arithmetic mean (\texttt{rank1}), and the PCA recommended weights
(\texttt{rank2}). The agreement of the three methods supports the use of
geometric mean as the dimension reduction method in the original index.
The scatterplot matrix also reveals the difference in ranking when the
three dimensions are given different weights in \texttt{rank3},
\texttt{rank4}, and \texttt{rank5}. The panel \texttt{rank4} vs
\texttt{rank5} shows a cluster of highlighted points with relatively
high ranks according to \texttt{rank5} but low in \texttt{rank4}. These
countries are printed in Table~\ref{tbl-hdi-expr} prints along with
their three dimensions, calculated indexes and rankings. While having a
high life expectancy and GNI \emph{per capita}, these countries
(Andorra, Qatar, San Marino, Kuwait, and Brunei Darussalam) score
relatively low in education, which is weighted heavily in
\texttt{index4}
(\texttt{0.1\ *\ life\_exp\ +\ 0.1\ *\ sch\ +\ 0.8\ *\ gni\_pc}).

{\centering \includegraphics{paper_files/figure-pdf/fig-hdi-expr-1.pdf}

\caption{\label{fig-hdi-expr}Scatterplot matrix of country rankings
computed from different dimension reduction expressions in HDI:
geometric mean (\texttt{rank0}), arithmetic mean (\texttt{rank1}),
principal component analysis recommended weights (\texttt{rank2}), and
heavier weight on life expectancy, education, and GNI per capita
(\texttt{rank3} - \texttt{rank5}). Geometric mean, arithmetic mean, and
PCA produce similar rankings where countries are aligned in a diagnoal
line, while country rankings varies when compare among \texttt{rank3} -
\texttt{rank5}. Panel \texttt{rank5} vs.~\texttt{rank4} highlights a
cluster of countries that rank high by \texttt{rank5} but low by
\texttt{rank4}. The plot demonstrates the necessity to tests on various
alternative expressions to study the robustness of the method and to
capture distinctive characteristics of different countries.}

\hypertarget{tbl-hdi-expr}{}
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.2118}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1882}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1176}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1765}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0824}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0824}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0706}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0706}}@{}}
\caption{\label{tbl-hdi-expr}A selected number of countries with low
rankings when education is given a heavy weight (\texttt{index4}) while
ranks high when a heavier weight is given on GNI per capital
(\texttt{index5}).}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
country
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
life expectancy
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
education
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
GNI per capita
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
index3
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
index4
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
rank3
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
rank4
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
country
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
life expectancy
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
education
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
GNI per capita
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
index3
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
index4
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
rank3
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
rank4
\end{minipage} \\
\midrule()
\endhead
Andorra & 0.929 & 0.721 & 0.942 & 0.909 & 0.764 & 32 & 69 \\
Qatar & 0.912 & 0.684 & 1.023 & 0.900 & 0.741 & 34 & 83 \\
San Marino & 0.937 & 0.701 & 0.947 & 0.914 & 0.749 & 31 & 76 \\
Kuwait & 0.903 & 0.670 & 0.947 & 0.884 & 0.721 & 39 & 96 \\
Brunei Darussalam & 0.841 & 0.694 & 0.977 & 0.840 & 0.737 & 60 & 85 \\
\bottomrule()
\end{longtable}

The maximum and minimum limit is carefully chosen with justification
provided in the technical notes:

\begin{quote}
The low minimum value for gross national income (GNI) per capita, \$100,
is justified by the considerable amount of unmeasured subsistence and
nonmarket production in economies close to the minimum, which is not
captured in the official data
\end{quote}

What if this lower minimum is raised or decreased?

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{res3 }\OtherTok{\textless{}{-}}\NormalTok{ res }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{swap\_values}\NormalTok{(}
    \AttributeTok{.module =}\NormalTok{ var\_trans, }\AttributeTok{.step =}\NormalTok{ rescale\_minmax, }\AttributeTok{.res =}\NormalTok{ gni\_pc,}
    \AttributeTok{.var =}\NormalTok{ min, }\AttributeTok{.values =} \FunctionTok{log10}\NormalTok{(}\DecValTok{700}\NormalTok{),}
    \AttributeTok{.raw\_data =}\NormalTok{ dt)}
\end{Highlighting}
\end{Shaded}

@@

\leavevmode\vadjust pre{\hypertarget{ref-UNDP_2021_tech_notes}{}}%
United Nations Development Programme. 2021. {``Human Development Report
2021-2022: The Next Frontier: Human Development and the Anthropocene.''}
\url{https://hdr.undp.org/sites/default/files/2021-22_HDR/hdr2021-22_technical_notes.pdf}.

\leavevmode\vadjust pre{\hypertarget{ref-hdi}{}}%
---------. 2022. {``Human Development Report 2021-22.''} New York.
\url{http://report.hdr.undp.org}.
